Subset coercions in \Coq

### 2006-05-04 papers/article/main.tex
Progress

\documentclass[twocolumn]{article}
\usepackage{abbrevs}
\usepackage{array}
%\usepackage{
\usepackage{amsmath}
\usepackage{amsthm}

@@

\newboolean{defineTheoremFr}
\newboolean{defineTheoremEn}
\setboolean{defineTheoremFr}{false}
\setboolean{defineTheoremEn}{true}
\input{mathenv}

@@

\author{Matthieu Sozeau \\
  LRI, Paris Sud XI University \\
  sozeau@lri.fr}
\title{Subset coercions in \Coq}
\date{\today}

@@

\section{\Russell{}}
The name of our language is an homage to the mathematician ? Russell who
discovered the famous paradox of set theory with the unrestricted
comprehension axiom. In this theory, it is possible to construct the set
$X = \{x `| x `; x\}$ whose definition is circular. Clearly, if $X `: X$
then $X `; X$ and if $X `; X$ then $X `: X$, hence we have an
inconsistency. \Russell{} was one of the pioneers of type theory when he
devised a set theory with a restricted comprehension axiom that
permitted to create subsets only from already defined subsets,
forbidding the definition of $X$.

@@

\subsection{From \ps{} to subset equivalence}
The \ps{} mechanism \cite{Rushby98:TSE, Shankar&Owre:WADT99}
is an extension of the \PVS type system which internalizes this idea.
Concretely, this means that the following rules are derivable in \PVS:
\begin{figure}[h]
  \begin{center}
    \AXC{$`G \vdash t : \{ x : T `| P[x] \}$}
    \UIC{$`G \vdash t : T$}
    \DP
    \AXC{$`G \vdash t : T$}
    \AXC{$`G \vdash P[t]$}
    \BIC{$`G \vdash t : \{ x : T `| P[x] \}$}
    \DP
  \end{center}
  \vspace{-1em}
  \label{PVS:rules}
  \caption{\PVS}
\end{figure}

@@

\subsection{Typing and conversion}
Our adaptation of \ps{} differs from the presentation of
given in \cite{Rushby98:TSE, Shankar&Owre:WADT99} essentially because of the
proof terms, but we have also extended it in some ways.

\subsubsection{The typing judgment}

\begin{figure*}[t]
    \def\seq{\typed}
    \def\fCenter{\wf}
    \begin{center}
      \WfEmpty\DP\WfVar\DP
      \Var\DP
    \end{center}

    \def\fCenter{\typed}
    \begin{center}
      \PropSet\DP
      \Subsum\DP
    \end{center}

    \Prod\DP
    \SigmaR\DP

    \vspace{\infvspace}
    \Abs\DP
    \SumDep\DP

    \vspace{\infvspace}
    \quad\App\DP
    \quad\quad
    \PiLeft\DP
    \quad
    \PiRight\DP

    \begin{center}\SubsetR\DP\end{center}
    \caption{\Russell{} typing}
  \label{fig:typing-decl-rules}
\end{figure*}

\Russell{}'s typing system (figure \vref{fig:typing-decl-rules}) is based
on the Calculus of Constructions \cite{coquandhuet88,Barras99}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is equivalent to $U$ in environment $`G$. 
Following the presentation as a pure type system \cite{barendregt91}, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$. As usual, we
let $s, s_i$ for $i `: `N$ range over sorts. The
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
set of rules $\PTSrules$ is defined by the following (functional) relation:
\begin{table}[h]
  \vspace{-1.5em}
  \axiomsd
  \vspace{-2em}
  \caption{Definition of $\PTSrules$}
  \label{fig:PTSrules}
\end{table}

@@

Likewise, we need to restrict the system so that it becomes property irrelevant for
the propositions appearing in subset types. Indeed, in \PVS{} you can't
manipulate proof terms, so you have
no way to create a term of type $0 \neq 0$ from the judgment $0 :
\mysubset{x}{`N}{P}$, but in a classical Calculus of Constructions with
sigma types or inductive types, the subset type is a first class object
defined with proper introduction and
elimination terms (figure \vref{fig:subset}). What we need to do is forbid the use of these terms
in \Russell{} terms, leaving only the type former.

@@

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \text{proj2\_sig}~t : P[\text{proj1\_sig}~t/x]$}
    \DP
  \end{center}
  \caption{Subset type in \CCI}
  \label{fig:subset}
\end{figure}

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also permits to include $\beta$-conversion
directly in the relation. The judgment $`G \typed T \subd{} U$ (figure
\vref{fig:conversion-decl-rules}) reads $T$ is convertible to $U$ in
environment $`G$. In the following, it is always assumed that the two
input types of the equivalence judgment are well-sorted, but maybe not by the same sort.

@@

\subsubsection{Properties}
\begin{paragraph}{Assumptions}
  We have proved basic metatheoretical results for this system in \Coq{}
  \cite{CCP}. We have assumed that it is strongly normalizing (SN) and has the
  following subject reduction (SR) property: if $`G \typed t : T$ and $t
  "->"_{\beta\pi} t'$ then $`G \typed t' : T$. Strong normalization
  could be proved using the generic proof method developed in
  \cite{geuvers95} but we leave it for future work. Gang Chen
  \cite{ChenPhD} has studied various type systems from the
  $\lambda$-cube extended by subtyping or coercive subtyping \cite{DBLP:conf/csl/Luo96},
  including the Calculus of Constructions, and proved such results as
  SN and SR for them. It requires a very careful analysis of the system 
  to avoid cycles in the proof due to the presence of dependent types and
  conversion ; hence we also leave it for future work.
\end{paragraph}

\begin{paragraph}{Notations}
  We write $`G \typed T, U : s$ for $`G \typed T : s$ and $`G \typed U :
  s$.
\end{paragraph}

As all functionnal PTS, \Russell{} has the unicity of sorting property,
hence we can prove:
\begin{lemma}[Well-sortedness of equivalence derivations]
  If $`G \typed T : s_1$, $`G \typed U : s_2$ and $`G \typed T \sub U$
  then $`G \typed T, U : s$ and $s = s_1 = s_2$.
\end{lemma}
\begin{proof}
  We will use the abbreviation $`G \typed T \sub U : s$ for $`G \typed T,
  U : s$ and $`G \typed T \sub U$ in the following.

  By induction on the subtyping derivation:
  \begin{list}{}
  \item[] \rulename{SubConv}: We have $T \eqbp U$, we apply unicity of sorting.

  \item[] \rulename{SubTrans}: We have by induction $`G \typed T \sub S : s$ and 
    $`G \typed S \sub U : s$. Hence we have $`G \typed T, U : s$.

  \item[] Other cases are straightforward.
  \end{list}
\end{proof}

@@

\subsubsection{Algorithmic typing}
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}

    \begin{center}
    \WfEmpty\DP

    \vspace{\infvspace}
    \WfVar\DP

    \vspace{\infvspace}
    \Var\DP

    \vspace{\infvspace}
    \PropSet\DP

@@

@@

    \SubsetR\DP
    \end{center}
  \caption{\Russell{} algorithmic typing and equivalence}
  \label{fig:typing-algo-rules}
\end{figure*}

@@

    \vspace{\infvspace}
    \SubProdA\DP
    \quad
    \SubSigmaA\DP

    \vspace{\infvspace}
    \SubProofA\DP
    \quad
    \SubSubAs\DP
  \end{center}
  \caption{\Russell{} algorithmic typing and equivalence}
  \label{fig:subtyping-algo-rules}
\end{figure*}

We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
$T `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
\ref{fig:mualgo}. In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator.

The next thing to do is to construct a decidable judgement for
equivalence given two well-sorted types. We denote it by $`G \typea T
\suba U$ (figure \ref{fig:subtyping-algo-rules}).
In our case, we have to check that conversion is decidable, which is the
case. We also need to eliminate the transitivity rule which is not
syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Elimination of transitivity]
  If $`G \typea T \suba U$ and $`G \typea U \suba V$, then $`G \typea T
  \suba V$.
\end{theorem}
\begin{proof}
  By lexicographic induction on the pair of derivations.
  \TODO{example}
\end{proof}

@@

Here is the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`E T, `G \typea t : T$ and $`G \typea T \suba U$ are decidable
  problems.
\end{theorem}

\subsection{From \CCq to \lng{}}
- Present \CCq
- Forgetful map

\section{From \lng{} to \CCq{}}
mutually recursive...

\subsection{Explicit coercions}
... derivation algo.

@@

\bibliography{bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{plain-url}

\renewcommand{\thefootnote}{}
\footnotetext{Ce rapport a été préparé sous \LaTeX~avec la fonte 
  \texttt{Concrete Math}}
\end{document}

>>>>

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}

\def\text{\mbox}

%\usepackage{amsmath}
%\usepackage{amsthm}

@@

\newboolean{defineTheoremFr}
\newboolean{defineTheoremEn}
\setboolean{defineTheoremFr}{false}
\setboolean{defineTheoremEn}{false}
\input{mathenv}

@@

\author{Matthieu Sozeau\inst{1}}

\institute{LRI, Paris Sud XI University\\\email{sozeau@lri.fr}}
\title{Subset coercions in \Coq}
\date{\today}

@@

\section{\Russell{}}
The name of our language is an homage to the mathematician Bertrand Russell who
discovered the famous paradox of set theory with the unrestricted
comprehension axiom. In this theory, it is possible to construct the set
$X = \{x `| x `; x\}$ whose definition is circular. Clearly, if $X `: X$
then $X `; X$ and if $X `; X$ then $X `: X$, hence we have an
inconsistency. \Russell{} was one of the pioneers of type theory when he
devised a set theory with a restricted comprehension axiom that
permitted to create subsets only from already defined subsets,
forbidding the definition of $X$ \cite{whitehead.russell:principia}.

@@

\subsection{From \ps{} to subset equivalence}
The \ps{} mechanism \cite{Rushby98:TSE,Shankar&Owre:WADT99}
is an extension of the \PVS type system which internalizes this idea.
Concretely, this means that the following rules are derivable in \PVS:
\begin{center}
  \AXC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \UIC{$`G \vdash t : T$}
  \DP
  \AXC{$`G \vdash t : T$}
  \AXC{$`G \vdash P[t]$}
  \BIC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \DP
\end{center}

@@

\subsection{Typing and conversion}
Our presentation of \ps{} differs slightly from the original one
essentially because of the proof terms, but we have also extended it in some ways.

\subsubsection{The typing judgment}
\begin{figure}[t]
  \begin{center}
    \def\infvspace{0.1in}
    \def\seq{\typed}
    \def\fCenter{\wf}

    \WfEmpty\DP\WfVar\DP

    \def\fCenter{\typed}
    \vspace{\infvspace}

    \Var\DP\PropSet\DP

    \vspace{\infvspace}
    \Subsum\DP

    \vspace{\infvspace}
    \App\DP

    \vspace{\infvspace}
    \SumDep\DP

    \vspace{\infvspace}
    \PiLeft\DP
    \quad
    \PiRight\DP

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} typing}
  \label{fig:typing-decl-rules}
  \vspace{-1em}
\end{figure}
%
\Russell{}'s typing system (figure \vref{fig:typing-decl-rules}) is based
on the Calculus of Constructions \cite{coquandhuet88,Barras99}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is equivalent to $U$ in environment $`G$. 
Following the presentation as a pure type system \cite{barendregt91}, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$. As usual, we
let $s, s_i$ for $i `: `N$ range over sorts. The
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
set of rules $\PTSrules$ is defined by the following (functional) relation:
% \begin{table}[h]
%   \vspace{-1.5em}
  \axiomsd
%   \vspace{-1em}
%   \caption{Definition of $\PTSrules$}
%   \label{fig:PTSrules}
% \end{table}

@@

Likewise, we need to restrict the system so that it becomes property irrelevant for
the propositions appearing in subset types. Indeed, in \PVS{} you can't
manipulate proof terms, so you have
no way to create a term of type $0 \neq 0$ from the judgment $0 :
\mysubset{x}{`N}{P}$, but in a classical Calculus of Constructions with
sigma types or inductive types, the subset type is a first class object
defined with proper introduction and
elimination terms (figure \vref{fig:cci-subset}). What we need to do is forbid the use of these terms
in \Russell{} terms, leaving only the type former.

@@

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \text{proj2\_sig}~t : P[\text{proj1\_sig}~t/x]$}
    \DP
  \end{center}
  \vspace{-1em}
  \caption{Subset type in \CCI}
  \label{fig:cci-subset}
  \vspace{-2.5em}
\end{figure}

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also conveys the idea that it can
include the usual $\beta\pi$-conversion directly in the judgement, 
although it may not be possible if we have cummulative universes.
The judgment $`G \typed T \subd{} U$ (figure
\vref{fig:conversion-decl-rules}  ) reads $T$ is convertible to $U$ in
environment $`G$. In the following, it is always assumed that the two
input types of the equivalence judgment are well-sorted, but maybe not by the same sort.

@@

\subsubsection{Properties}
\begin{paragraph}{Assumptions}
  We have proved basic metatheoretical results for this system in \Coq{}
  \cite{coq/Russell/meta}. We have assumed that it is strongly normalizing (SN) and has the
  following subject reduction (SR) property: if $`G \typed t : T$ and $t
  "->"_{\beta\pi} t'$ then $`G \typed t' : T$. Gang Chen
  \cite{ChenPhD} has studied various type systems from the
  $\lambda$-cube extended by subtyping or coercive subtyping \cite{DBLP:conf/csl/Luo96},
  including the Calculus of Constructions, and proved such results as
  SN and SR for them. It requires a very careful analysis of the system 
  to avoid cycles in the proof due to the presence of dependent types and
  conversion ; hence we leave it for future work. Strong normalization
  could also be proved, perhaps more easily, using the generic proof method developed in
  \cite{geuvers95}, by finding a proper interpretation of \Russell terms
  into the Calculus of Constructions.
\end{paragraph}

As all functionnal PTS, \Russell{} has the unicity of sorting property,
hence we can prove:
\begin{lemma}[Well-sortedness of equivalence derivations]
  If $`G \typed T : s_1$, $`G \typed U : s_2$ and $`G \typed T \sub U$
  then $`G \typed T, U : s$\footnote{We write $`G \typed T, U : s$ for $`G \typed T : s$ and $`G \typed U :
  s$.} and $s = s_1 = s_2$.
\end{lemma}
\begin{proof}
  We will use the abbreviation $`G \typed T \sub U : s$ for $`G \typed T,
  U : s$ and $`G \typed T \sub U$ in the following.

  By induction on the subtyping derivation:
  \begin{list}{-}
  \item \rulename{SubConv}: We have $T \eqbp U$, we apply unicity of sorting.

  \item[-] \rulename{SubTrans}: We have by induction $`G \typed T \sub S : s$ and 
    $`G \typed S \sub U : s$. Hence we have $`G \typed T, U : s$.

  \item Other cases are straightforward.
  \end{list}
\end{proof}

@@

\subsubsection{Algorithmic typing}
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

%     \vspace{\infvspace}
%     \WfVar\DP

%     \vspace{\infvspace}
%     \Var\DP

%     \vspace{\infvspace}
%     \PropSet\DP

%     \vspace{\infvspace}
%     \Prod\DP

%     \vspace{\infvspace}
%     \Abs\DP

@@

%     \vspace{\infvspace}
%     \SigmaR\DP

@@

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} algorithmic typing, new rules}
  \label{fig:typing-algo-rules}
  \vspace{-1em}
\end{figure*}

@@

    \vspace{\infvspace}
    \SubProdA\DP

    \vspace{\infvspace}
    \SubSigmaA\DP

    \vspace{\infvspace}
    \SubProofA\DP

    \vspace{\infvspace}
    \SubSubAs\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} algorithmic equivalence}
  \label{fig:subtyping-algo-rules}
  \vspace{-1em}
\end{figure*}

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
$T `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
\ref{fig:mualgo}. In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator.
\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two well-sorted types. We denote it by $`G \typea T
  \suba U$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity rule which is not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Elimination of transitivity]
  If $`G \typea T \suba U$ and $`G \typea U \suba V$, then $`G \typea T
  \suba V$.
\end{theorem}
\begin{proof}
  By lexicographic induction on the pair of derivations.
  \TODO{example}
\end{proof}
\end{paragraph}

@@

Finaly, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U, `G \typea T \suba U$ are decidable
  problems.
\end{theorem}

\subsection{From \CC to \lng{}}
We have presented a calculus based on \CC with a stronger equivalence
but with restricted typing rules. We will now see how they relate
formally.
We can build a forgetful map from terms of \CCI to \Russell
(interpreting inductives as constants). It bears
relation to the $`e$ extraction function defined in
\cite{Werner:ProofIrrelevance}. Essentially, \Russell{} terms do not
contain or manipulate logical information attached to objects of subset
types, hence we must forget about it when translating.

We define the forgetful map $()^{\circ}$ as an homomorphism on terms
except for the following cases:
\begin{center}
  \CCtoRussellforgetFig
\end{center}

We can now prove the following:
\begin{theorem}[Forgetful map correctness]
  If $`G \typec t : T$ then $`G^{\circ} \typea t^{\circ} : T^{\circ}$ if $()^{\circ}$ is
  defined on $`G, t$ and $T$.
\end{theorem}

Practically, this means we can use almost any existing definition in the
\Coq environment, potentially turning them (locally) to an opaque form
if their body is not traductible. Obviously, this map is involutive and
the identity on \Russell terms.

\section{From \lng{} to \CCq{}}
We will now build an interpretation
$\ip{t}{`G}$ of \Russell terms $t$ in environment $`G$ into \CCI
terms. We will check at the end that it respects the $()^{\circ}$ operator
in the sense that if $t$ is well-typed in $`G$ then $\ip{t}{`G}^{\circ} = t$.

Our interpretation will build a full-fledged \CCI term from its
algorithmic skeleton and a rich type written in \Russell. Obviously, we
cannot infer proof terms where they are needed in \CCI, but we can build
a \emph{partial} term, leaving typed holes where proofs are
needed. Hence we add a rule to build existential variables (or
metavariables) to the target calculus ($\typeq$ denotes the new system's typing judgement):
\begin{prooftree}
  \UAX{}
  {$`G \typeq P : \Prop$}
  {$`G \typeq ?_{P} : P$}
  {}
\end{prooftree}

We restrict it to objects of type $\Prop$ because we consider that the
informative part of the algorithm has been entirely given in the
original term.

Now we can build a total interpretation:
\begin{proposition}[Interpretation correctness]
  If $`G \typea t : T$ then $\ipG{`G} \typeq \ip{t}{`G} : \ip{T}{`G}$.
\end{proposition}

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organized as follows: first we will
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we will define the
interpretation of terms which is mutually recursive with the first
one. Finaly we will present a proof sketch of the aforementionned
proposition.

\subsection{Explicit coercions}
The algorithm which derives explicit coercions is based on the
algorithmic equivalence derivations.

@@

\bibliography{bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{splncs}

\renewcommand{\thefootnote}{}
\footnotetext{This article was typeset in \LaTeX~using the
  \texttt{Concrete Math} font}
\end{document}
